---
title: "1870 Census Occupation"
author: "Alusine Samura"
date: "2025-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task Description

In this task, we will be working with the 1870 Census Data on people's occupation, which contains state-level aggregates of occupation by gender. The goal is to:

- Use `tidyr` to reshape the data into a long format.
- Separate the `occupation.gender` variable into two variables: `occupation` and `gender`.
- Create scatter plots to compare the values for men against women, faceted by occupation.

```{r}
# Load necessary libraries
library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)  # Updated to use read_csv()

# Load the dataset directly from the URL (or use your local path)
data <- read_csv("occupation-1870.csv")

# Check the first few rows of the data to understand its structure
head(data)

# Dynamically select columns corresponding to male and female counts
cols <- grep("Male|Female", colnames(data), value = TRUE)

# Gather the data into long format using pivot_longer
data_long <- data %>%
  pivot_longer(cols = all_of(cols), names_to = "occupation.gender", values_to = "count")

# Separate the "occupation.gender" column into occupation and gender
data_long <- data_long %>%
  separate(occupation.gender, into = c("occupation", "gender"), sep = "\\.")

# Check data after separation
head(data_long)

# Pivot the data to wide format (one column per gender)
data_wide <- data_long %>%
  pivot_wider(names_from = "gender", values_from = "count", values_fill = list(count = 0))

# Check the wide-format data
head(data_wide)

# Create a scatter plot comparing men's and women's counts, facetted by occupation
ggplot(data_wide, aes(x = Male, y = Female, color = occupation)) +
  geom_point(size = 3, alpha = 0.7) +  # Adjust size and transparency of points
  facet_wrap(~ occupation) +
  labs(
    title = "Occupation Comparison by Gender in the 1870 Census",
    subtitle = "Men vs. Women by Occupation",
    x = "Men's Count",
    y = "Women's Count",
    caption = "Data from the 1870 Census"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "italic"),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "italic")
  )

```

